---
title: "Conservation Biogeography of *Pongo pygmaeus*: A 25-Year GBIF Data Synthesis"
subtitle: "Spatial and Temporal Dynamics of Bornean Orangutan Occurrence Records (2000-2024)"
author:
  - name: "Jarian Permana"
    affiliation: "Independent Researcher"
    email: "jarian.learning@gmail.com"
    orcid: "0000-0000-0000-0000"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    theme:
      light: journal
      dark: darkly
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "📑 Table of Contents"
    code-fold: true
    code-tools:
      source: true
      toggle: false
    df-print: paged
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
    embed-resources: true
    highlight-style: tango
    mainfont: "Arial"
    fontsize: "1.1rem"
    linkcolor: "#2E8B57"
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
editor: source
bibliography: references.bib

---

```{r setup, include=FALSE}
# Setup chunk - Load all necessary libraries and load saved environment
library(htmltools)
library(DiversityStats)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(DT)
library(plotly)
library(leaflet)
library(knitr)
library(kableExtra)
library(here)
library(scales)
library(RColorBrewer)

# =========================================================================
# 1. ENVIRONMENT CONFIGURATION
# =========================================================================

# Suppress package startup messages
options(
  dplyr.summarise.inform = FALSE,
  readr.show_progress = FALSE,
  warnings.length = 200
)

# Load libraries with error handling
required_packages <- c(
  "dplyr", "tidyr", "ggplot2", "sf", "rnaturalearth", "rnaturalearthdata",
  "DT", "plotly", "leaflet", "knitr", "kableExtra", "here", "scales", 
  "RColorBrewer", "htmltools"
)

invisible(lapply(required_packages, function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    stop(paste("Package", pkg, "is required but not installed"))
  }
}))

# Define color palettes (Nature-inspired)
COLORS <- list(
  subspecies = c(
    "morio" = "#D55E00",      # Burnt orange - NE Borneo
    "pygmaeus" = "#0072B2",    # Deep blue - NW Borneo  
    "wurmbii" = "#009E73",      # Teal green - SW Borneo
    "Unidentified" = "#999999"  # Gray
  ),
  countries = c(
    "Indonesia" = "#2E8B57",    # Sea green
    "Malaysia" = "#FFD700"       # Gold
  ),
  temporal = c(
    "2000-2004" = "#FDE725",
    "2005-2009" = "#B8DE29",
    "2010-2014" = "#6DCD59",
    "2015-2019" = "#20A486",
    "2020-2024" = "#31688E"
  )
)

# Load and validate data
data_path <- here::here("data", "envData.RData")
if (!file.exists(data_path)) {
  stop("Data file not found. Please ensure 'data/envData.RData' exists.")
}

load(data_path)

# Standardize object name
pongo_data <- if (exists("pongo_clean")) {
  pongo_clean
} else if (exists("pongo_data")) {
  pongo_data
} else {
  stop("No valid data object found in envData.RData")
}

# Data validation
required_columns <- c("decimalLongitude", "decimalLatitude", "year", 
                      "subspecies", "country_name", "stateProvince")
missing_cols <- required_columns[!required_columns %in% names(pongo_data)]
if (length(missing_cols) > 0) {
  warning(paste("Missing columns:", paste(missing_cols, collapse = ", ")))
}

# Data quality report
data_quality <- list(
  total_records = nrow(pongo_data),
  complete_coordinates = sum(!is.na(pongo_data$decimalLongitude) & 
                              !is.na(pongo_data$decimalLatitude)),
  complete_years = sum(!is.na(pongo_data$year)),
  unique_institutions = n_distinct(pongo_data$institutionCode, na.rm = TRUE),
  unique_locations = n_distinct(pongo_data$stateProvince, na.rm = TRUE),
  identified_subspecies = sum(pongo_data$subspecies != "Unidentified", na.rm = TRUE)
)

# Set seed for reproducibility
set.seed(2024)

# Prepare spatial data with error handling
tryCatch({
  world <- ne_countries(scale = "medium", returnclass = "sf")
  borneo_bbox <- c(xmin = 108, xmax = 120, ymin = -4, ymax = 8)
  borneo_map <- st_crop(world, borneo_bbox)
  rm(world)
}, error = function(e) {
  warning("Could not load natural earth data. Using simplified geometry.")
  borneo_map <<- NULL
})

# Create figure directory if needed
fig_dir <- here::here("outputs", "figures")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
```

# Abstract 

Background: The Bornean orangutan (Pongo pygmaeus) is a critically endangered great ape endemic to the island of Borneo. Understanding its distribution patterns is essential for effective conservation planning.

Methods: This study analyzes 24 years (2000-2024) of occurrence records from the Global Biodiversity Information Facility (GBIF), focusing on Indonesia and Malaysia. Data were filtered for coordinate completeness and geospatial quality, then analyzed for spatial, temporal, and taxonomic patterns.

Results: The analysis reveals significant temporal trends in occurrence reporting and distinct spatial partitioning among the three recognized subspecies: *P. p. morio, P. p. pygmaeus, and P. p. wurmbii*.

Conclusions: These findings provide crucial baseline data for orangutan conservation and highlight priority areas for future survey efforts.


```{r keyabstract, include=TRUE, echo=FALSE, results='asis'}
# Key metrics box for abstract
cat('<div style="background-color: #f0f7f0; border-left: 4px solid #2E8B57; padding: 15px; margin: 20px 0; border-radius: 5px;">')
cat('<strong>📊 Key metrics:</strong> ')
cat(format(data_quality$total_records, big.mark = ","))
cat(" occurrence records analyzed, ")
cat(data_quality$unique_locations)
cat(" unique locations, ")
cat(round(data_quality$identified_subspecies/data_quality$total_records*100, 1))
cat("% identified to subspecies level.")
cat('</div>')
```

# 1. Introduction

The Bornean orangutan (*Pongo pygmaeus* Linnaeus, 1760) represents one of our closest living relatives and a flagship species for Southeast Asian biodiversity conservation [@delgado2008]. Classified as Critically Endangered on the IUCN Red List, the species has experienced dramatic population declines due to habitat loss, fragmentation, and hunting [@singh2022].

## 1.1 Research Objectives This study aims to:

1. **Quantify** the spatial distribution of *P. pygmaeus* occurrence records across Borneo
2. **Analyze** temporal trends in data collection and reporting (2000-2024)
3. **Compare** occurrence patterns among the three recognized subspecies
4. **Evaluate** data completeness and quality from GBIF repositories

## 1.2 Subspecies Overview

| ~Subspecies~ | Common Name | ~Geographic Range~ | ~Primary Locations~ |
|------------------|------------------|------------------|------------------|
| ~*P. p. morio*~ | ~Northeast Bornean~ | ~Sabah, East Kalimantan~ | ~Kinabatangan, Danum Valley~ |
| ~*P. p. pygmaeus*~ | ~Northwest Bornean~ | ~Sarawak, West Kalimantan~ | ~Batang Ai, Gunung Gading~ |
| ~*P. p. wurmbii*~ | ~Southwest Bornean~ | ~Central & South Kalimantan~ | ~Tanjung Puting, Gunung Palung~ |

# 2. Executive Summary

## 2.1 Key Findings
```{r KeyFinding, include=TRUE, echo=FALSE, results='asis'}
# Create professional summary cards - FORMAT YANG BENAR
cat('<div style="display: flex; justify-content: space-between; flex-wrap: wrap; gap: 15px; margin: 25px 0;">')

metrics <- list(
  list(
    icon = "🌿",
    color = "#2E8B57",
    value = format(data_quality$total_records, big.mark = ","),
    label = "Total Occurrence Records"
  ),
  list(
    icon = "📅",
    color = "#377EB8",
    value = "2000-2024",
    label = "Time Period"
  ),
  list(
    icon = "📍",
    color = "#D55E00",
    value = data_quality$unique_locations,
    label = "Unique Locations"
  ),
  list(
    icon = "🏛️",
    color = "#FFD700",
    value = data_quality$unique_institutions,
    label = "Contributing Institutions"
  ),
  list(
    icon = "🧬",
    color = "#009E73",
    value = paste0(round(data_quality$identified_subspecies/data_quality$total_records*100, 1), "%"),
    label = "Subspecies Identified"
  )
)

for (m in metrics) {
  cat(sprintf('
<div style="background-color: %s; color: white; padding: 15px; border-radius: 10px; text-align: center; flex: 1; min-width: 150px; margin: 5px; box-shadow: 0 2px 4px rgba(0,0,0,0.1);">
  <div style="font-size: 2rem;">%s</div>
  <div style="font-size: 1.5rem; font-weight: bold; margin: 8px 0;">%s</div>
  <div style="font-size: 0.8rem; opacity: 0.9;">%s</div>
</div>', m$color, m$icon, m$value, m$label))
}

cat('</div>')
```

## 2.2 Data Quality Assessment

```{r DataQuality, include=TRUE}
quality_df <- data.frame(
  Metric = c(
    "Total occurrence records",
    "Records with complete coordinates",
    "Records with complete temporal data",
    "Unique contributing institutions",
    "Unique geographic locations",
    "Records with subspecies identification"
  ),
  Value = c(
    format(data_quality$total_records, big.mark = ","),
    paste0(format(data_quality$complete_coordinates, big.mark = ","), 
           " (", round(data_quality$complete_coordinates/data_quality$total_records*100, 1), "%)"),
    paste0(format(data_quality$complete_years, big.mark = ","),
           " (", round(data_quality$complete_years/data_quality$total_records*100, 1), "%)"),
    data_quality$unique_institutions,
    data_quality$unique_locations,
    paste0(format(data_quality$identified_subspecies, big.mark = ","),
           " (", round(data_quality$identified_subspecies/data_quality$total_records*100, 1), "%)")
  )
)

kable(quality_df, format = "html", 
      caption = "Table 1: Data Quality Assessment Summary",
      align = "lc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, font_size = 14) %>%
  row_spec(0, background = "#2E8B57", color = "white")
```

# 3. Materials and Methods

## 3.1 Data Acquisition

Occurrence data were accessed through the GBIF API using the rgbif R package [@chamberlain2021]. The query parameters were:

```{r DataAcq, include=TRUE, results='asis'}
# Create table with proper formatting
# Create table with proper formatting
cat('
<div style="margin: 20px 0;">
<table class="table table-striped table-hover" style="width: 80%; margin: auto;">
  <thead>
    <tr style="background-color: #2E8B57; color: white;">
      <th style="text-align: left; padding: 12px;">Parameter</th>
      <th style="text-align: left; padding: 12px;">Value</th>
      <th style="text-align: left; padding: 12px;">Justification</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td style="padding: 10px;"><strong>Taxon</strong></td>
      <td style="padding: 10px;"><em>Pongo pygmaeus</em></td>
      <td style="padding: 10px;">Target species for conservation analysis</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Year range</strong></td>
      <td style="padding: 10px;">2000-2024</td>
      <td style="padding: 10px;">Recent data focus (25-year period)</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Coordinate status</strong></td>
      <td style="padding: 10px;">Present</td>
      <td style="padding: 10px;">Required for spatial analysis</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Geospatial issues</strong></td>
      <td style="padding: 10px;">Excluded</td>
      <td style="padding: 10px;">Ensure data quality</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Occurrence status</strong></td>
      <td style="padding: 10px;">PRESENT</td>
      <td style="padding: 10px;">Include only valid records</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Geographic focus</strong></td>
      <td style="padding: 10px;">Indonesia (ID) & Malaysia (MY)</td>
      <td style="padding: 10px;">Bornean range only</td>
    </tr>
  </tbody>
</table>
</div>
')
```

## 3.2 Data Processing Pipeline

The analytical workflow consisted of six sequential steps

```{r DataPipe, include=TRUE, fig.height=4, fig.width=12}
# Enhanced workflow diagram
workflow_data <- data.frame(
  step = c("1. GBIF API\nData Download", 
           "2. Data Filtering\n(ID & MY only)", 
           "3. Data Cleaning\n(Coordinates, Issues)", 
           "4. Data Enrichment\n(Subspecies, Year groups)", 
           "5. Statistical\nAnalysis", 
           "6. Visualization\n& Reporting"),
  x = 1:6,
  y = 1,
  description = c("Raw occurrence data from GBIF",
                  "Keep Indonesia & Malaysia records",
                  "Remove records with issues",
                  "Add subspecies & temporal groups",
                  "Spatial & temporal analysis",
                  "Generate plots and tables")
)

# Create enhanced workflow plot
workflow_plot <- ggplot(workflow_data, aes(x = x, y = y)) +
  # Background bars
  geom_rect(aes(xmin = x - 0.4, xmax = x + 0.4, 
                ymin = y - 0.3, ymax = y + 0.3),
            fill = "#2E8B57", alpha = 0.15, color = "#2E8B57", linewidth = 0.5) +
  # Step numbers
  geom_point(size = 12, color = "#2E8B57", alpha = 0.3) +
  geom_text(aes(label = x), size = 5, fontface = "bold", color = "#2E8B57") +
  # Step titles
  geom_text(aes(label = step, y = y - 0.5), size = 3.5, fontface = "bold") +
  # Descriptions
  geom_text(aes(label = description, y = y - 0.9), size = 2.8, color = "gray40") +
  # Arrows between steps
  geom_segment(aes(x = x + 0.4, xend = x + 0.6, y = y, yend = y),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               color = "#2E8B57", size = 0.8, data = workflow_data[1:5,]) +
  theme_void() +
  xlim(0.5, 6.5) +
  ylim(0.2, 1.5) +
  ggtitle("Data Processing Workflow") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

workflow_plot
```

## 3.3 Statistical Analysis

- Analyses were conducted using R version 4.3 with the following approaches:

- Temporal analysis: Histogram decomposition with 5-year rolling windows

- Spatial analysis: Point pattern visualization with subspecies stratification

- Comparative analysis: Country-level and location-level aggregations

# 4. Results

## 4.1 Spatial Distribution Patterns

```{r SpatDist, include=TRUE}
# Prepare sampling data
map_data <- pongo_data %>%
  filter(!is.na(decimalLongitude) & !is.na(decimalLatitude))

if(nrow(map_data) > 5000) {
  set.seed(2024)
  map_data <- map_data %>% sample_n(5000)
}

# Create static map
p_distribution <- ggplot() +
  {if(!is.null(borneo_map)) geom_sf(data = borneo_map, fill = "#F5F5DC", color = "#333333", linewidth = 0.3)} +
  geom_point(data = map_data,
             aes(x = decimalLongitude, y = decimalLatitude, 
                 color = subspecies, shape = subspecies),
             alpha = 0.6, size = 1.8) +
  coord_sf(xlim = c(108, 120), ylim = c(-4, 8)) +
  theme_minimal(base_size = 12) +
  scale_color_manual(name = "Subspecies", values = COLORS$subspecies) +
  scale_shape_manual(name = "Subspecies", 
                     values = c("morio" = 16, "pygmaeus" = 17, 
                               "wurmbii" = 15, "Unidentified" = 4)) +
  labs(title = "Figure 1: Geographic Distribution of Pongo pygmaeus Occurrence Records",
       subtitle = paste("Borneo, 2000-2024 (n =", format(nrow(map_data), big.mark = ","), "points)"),
       x = "Longitude (°E)", y = "Latitude (°N)",
       caption = "Data source: GBIF (2024)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        legend.position = "right",
        panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = NA, color = "#333333", linewidth = 0.5))

p_distribution
```

## 4.2 Temporal Trends

```{r TempTrend, include= TRUE}
# Jika tetap ingin menggunakan LOESS
annual_counts <- pongo_data %>%
  group_by(year) %>%
  summarise(count = n(), .groups = 'drop') %>%
  complete(year = 2000:2024, fill = list(count = 0))

# Check if there are enough points for LOESS
if(nrow(annual_counts) >= 10) {
  p_annual <- ggplot(annual_counts, aes(x = year, y = count)) +
    geom_col(fill = "#377EB8", color = "white", alpha = 0.7, width = 0.8) +
    geom_smooth(method = "loess", se = TRUE, color = "#D55E00", 
                fill = "#D55E00", alpha = 0.2, size = 1.2) +
    theme_minimal(base_size = 12) +
    labs(title = "Figure 2: Temporal Trends in Occurrence Reporting",
         subtitle = "Annual record counts with LOESS smoothing",
         x = "Year", y = "Number of Occurrence Records",
         caption = "Shaded area represents 95% confidence interval") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
    scale_y_continuous(labels = label_comma())
  
  p_annual
} else {
  # Fallback to linear model if insufficient data
  p_annual <- ggplot(annual_counts, aes(x = year, y = count)) +
    geom_col(fill = "#377EB8", color = "white", alpha = 0.7, width = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "#D55E00", 
                fill = "#D55E00", alpha = 0.2, size = 1.2) +
    theme_minimal(base_size = 12) +
    labs(title = "Figure 2: Temporal Trends in Occurrence Reporting",
         subtitle = "Annual record counts with linear regression (insufficient data for LOESS)",
         x = "Year", y = "Number of Occurrence Records") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
    scale_y_continuous(labels = label_comma())
  
  p_annual
}
```

## 4.3 Subspecies Partitioning

```{r SubPart, include= TRUE}
# Stacked bar chart by period
subspecies_period <- pongo_data %>%
  filter(subspecies != "Unidentified") %>%
  group_by(year_group, subspecies) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(year_group) %>%
  mutate(prop = count / sum(count) * 100)

p_subspecies_stack <- ggplot(subspecies_period, 
                              aes(x = year_group, y = count, fill = subspecies)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7) +
  theme_minimal(base_size = 12) +
  scale_fill_manual(name = "Subspecies", values = COLORS$subspecies) +
  scale_y_continuous(labels = scales::percent, name = "Proportion (%)") +
  labs(title = "Figure 3: Subspecies Composition Through Time",
       subtitle = "Relative proportion by 5-year period",
       x = "Time Period", y = "Proportion of Records",
       caption = "Excludes unidentified specimens") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  geom_text(aes(label = paste0(round(prop, 1), "%")), 
            position = position_fill(vjust = 0.5), size = 3)

p_subspecies_stack
```

## 4.4 Country-Level Comparison

```{r CountryComp, include= TRUE}
p_country_enhanced <- pongo_data %>%
  group_by(year_group, country_name) %>%
  summarise(count = n(), .groups = 'drop') %>%
  ggplot(aes(x = year_group, y = count, fill = country_name, 
             group = country_name)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = format(count, big.mark = ",")), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 3) +
  theme_minimal(base_size = 12) +
  scale_fill_manual(name = "Country", values = COLORS$countries) +
  labs(title = "Figure 4: Country-Level Comparison of Occurrence Records",
       subtitle = "Indonesia vs. Malaysia, 2000-2024",
       x = "Time Period", y = "Number of Occurrence Records",
       caption = "Records from Bornean territories only") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

p_country_enhanced
```

## 4.5 Seasonal Patterns

```{r SeasonPatt, include=TRUE}
heatmap_data <- pongo_data %>%
  filter(!is.na(month) & !is.na(year)) %>%
  group_by(year, month) %>%
  summarise(count = n(), .groups = 'drop')

if(nrow(heatmap_data) > 0) {
  p_seasonal <- ggplot(heatmap_data, aes(x = year, y = month, fill = count)) +
    geom_tile() +
    theme_minimal(base_size = 11) +
    scale_fill_gradientn(name = "Record\nCount", 
                         colors = c("#FDE725", "#B8DE29", "#6DCD59", "#20A486", "#31688E"),
                         trans = "log10",
                         labels = label_comma()) +
    scale_y_continuous(breaks = 1:12, labels = month.abb, name = "Month") +
    scale_x_continuous(breaks = seq(2000, 2024, by = 4), name = "Year") +
    labs(title = "Figure 5: Seasonal Distribution of Occurrence Records",
         subtitle = "Heatmap of record counts by year and month",
         caption = "Color scale represents log10-transformed counts") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5),
          legend.position = "right")
  
  p_seasonal
} else {
  cat("*Insufficient month/year data for seasonal analysis*")
}
```

## 4.6 Top Contributing Locations

```{r TopCont, include=TRUE}
top_10_locations <- pongo_data %>%
  filter(stateProvince != "Unidentified Province") %>%
  group_by(stateProvince, country_name) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count)) %>%
  head(10)

p_top_locations <- ggplot(top_10_locations, 
                          aes(x = reorder(stateProvince, count), 
                              y = count, fill = country_name)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  scale_fill_manual(name = "Country", values = COLORS$countries) +
  labs(title = "Figure 6: Top 10 Sampling Locations",
       subtitle = "Provinces/States with highest occurrence records",
       x = "Location (Province/State)", y = "Number of Occurrence Records",
       caption = "Excludes unidentified locations") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "top") +
  geom_text(aes(label = format(count, big.mark = ",")), 
            hjust = -0.2, size = 3.5)

p_top_locations
```

# 5. Discussion

## 5.1 Spatial Patterns

The distribution maps reveal clear spatial partitioning among the three subspecies, consistent with published range descriptions [@ancrenaz2016]. P. p. wurmbii dominates records from southern Borneo, while P. p. morio and P. p. pygmaeus show stronghold in northeastern and northwestern regions respectively.

## 5.2 Temporal Trends

The temporal analysis reveals:

1.  Exponential growth in occurrence records since 2015 (+300% from 2010-2014 baseline)

2.  Sampling bias toward accessible areas (roads, research stations)

3.  Underrepresentation of interior, mountainous regions

## 5.3 Data Gaps and Limitations

```{r DatLimit, include=TRUE, echo=FALSE,results='asis'}
cat('
<div style="margin: 20px 0;">
<table class="table table-striped table-hover" style="width: 100%; margin: auto;">
  <thead>
    <tr style="background-color: #D55E00; color: white;">
      <th style="text-align: left; padding: 12px;">Limitation</th>
      <th style="text-align: left; padding: 12px;">Impact on Analysis</th>
      <th style="text-align: left; padding: 12px;">Mitigation Strategy</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td style="padding: 10px;"><strong>Sampling Bias</strong></td>
      <td style="padding: 10px;">Overrepresentation of accessible areas (roads, research stations, protected areas)</td>
      <td style="padding: 10px;">Acknowledge in interpretations; focus on relative patterns</td>
    </tr>
    <tr style="background-color: #f9f9f9;">
      <td style="padding: 10px;"><strong>Taxonomic Uncertainty</strong></td>
      <td style="padding: 10px;">~30% of records not identified to subspecies level</td>
      <td style="padding: 10px;">Conservative subspecies assignments; exclude unidentified from subspecies-specific analyses</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Temporal Clustering</strong></td>
      <td style="padding: 10px;">Recent data (2015-2024) dominate trends</td>
      <td style="padding: 10px;">Use 5-year rolling windows; analyze proportional trends</td>
    </tr>
    <tr style="background-color: #f9f9f9;">
      <td style="padding: 10px;"><strong>Coordinate Uncertainty</strong></td>
      <td style="padding: 10px;">Some records have imprecise coordinates</td>
      <td style="padding: 10px;">Filter for high-quality coordinates; use GBIF geospatial filters</td>
    </tr>
    <tr>
      <td style="padding: 10px;"><strong>Institutional Bias</strong></td>
      <td style="padding: 10px;">Most data from few major institutions</td>
      <td style="padding: 10px;">Identify data gaps; encourage more institutional participation</td>
    </tr>
    <tr style="background-color: #f9f9f9;">
      <td style="padding: 10px;"><strong>Geographic Gaps</strong></td>
      <td style="padding: 10px;">Underrepresentation of interior, mountainous, and border regions</td>
      <td style="padding: 10px;">Priority areas for future surveys</td>
    </tr>
  </tbody>
</table>
</div>
')
```


# 6. Conclusions and Recommendations

## 6.1 Key Conclusions
1. GBIF provides valuable occurrence data for Pongo pygmaeus with 24+ years of temporal coverage

2. Subspecies show distinct spatial partitioning that aligns with published biogeographic boundaries

3. Recent years (2015-2024) show dramatic increase in reporting, indicating growing research attention

4. Indonesia contributes the majority (62%) of occurrence records


## 6.2 Conservation Recommendations
Based on these findings, we recommend:

1. Targeted surveys in under-sampled regions (Central Kalimantan interior, Sarawak highlands)

2. Standardized subspecies identification protocols for field researchers

3. Long-term monitoring sites in each subspecies' core range

4. Data sharing incentives for local institutions and citizen scientists


# 7. Interactive Data Explorer
## 7.1 Interactive Distribution Map

```{r IntMap, include=TRUE}
# Prepare interactive map data
interactive_data <- pongo_data %>%
  filter(!is.na(decimalLongitude) & !is.na(decimalLatitude)) %>%
  slice_sample(n = min(nrow(.), 3000))

leaflet(interactive_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = 114, lat = 2, zoom = 6) %>%
  addCircleMarkers(
    lng = ~decimalLongitude,
    lat = ~decimalLatitude,
    radius = 3,
    color = ~COLORS$subspecies[subspecies],
    fillOpacity = 0.5,
    opacity = 0.7,
    popup = ~paste(
      "<strong>Subspecies:</strong>", subspecies, "<br>",
      "<strong>Location:</strong>", stateProvince, "<br>",
      "<strong>Year:</strong>", year, "<br>",
      "<strong>Institution:</strong>", institutionCode
    ),
    label = ~paste(subspecies, "-", stateProvince),
    group = ~subspecies
  ) %>%
  addLayersControl(
    overlayGroups = unique(interactive_data$subspecies),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomright",
    colors = COLORS$subspecies,
    labels = names(COLORS$subspecies),
    title = "Subspecies"
  )
```


## 7.2 Interactive Data Table
```{r IntDatTable, include=TRUE}
DT::datatable(
  pongo_data %>%
    select(gbifID, year, subspecies, country_name, stateProvince, 
           institutionCode, decimalLongitude, decimalLatitude) %>%
    rename(
      ID = gbifID,
      Year = year,
      Subspecies = subspecies,
      Country = country_name,
      Province = stateProvince,
      Institution = institutionCode,
      Longitude = decimalLongitude,
      Latitude = decimalLatitude
    ),
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf')
  ),
  filter = 'top',
  rownames = FALSE,
  class = 'compact hover',
  extensions = 'Buttons'
) %>%
  formatStyle(
    columns = 'Subspecies',
    backgroundColor = styleEqual(
      names(COLORS$subspecies),
      unname(COLORS$subspecies)
    ),
    color = styleEqual(
      names(COLORS$subspecies),
      ifelse(names(COLORS$subspecies) == "Unidentified", "black", "white")
    )
  )
```


# 8. Supplementary Materials
## 8.1 Complete Statistical Tables
```{r CompStat, include=TRUE}
# Table S1: Statistics by subspecies
stats_subspecies <- pongo_data %>%
  filter(subspecies != "Unidentified") %>%
  group_by(subspecies) %>%
  summarise(
    `Total Records` = format(n(), big.mark = ","),
    `Percentage` = paste0(round(n()/data_quality$identified_subspecies * 100, 1), "%"),
    `Year Range` = paste(min(year, na.rm = TRUE), "-", max(year, na.rm = TRUE)),
    `Unique Locations` = n_distinct(stateProvince, na.rm = TRUE),
    `Unique Institutions` = n_distinct(institutionCode, na.rm = TRUE)
  )

kable(stats_subspecies, format = "html", 
      caption = "Table S1: Subspecies-Level Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Table S2: Complete location list
location_stats <- pongo_data %>%
  group_by(stateProvince, country_name) %>%
  summarise(
    Records = n(),
    Subspecies_present = paste(unique(subspecies[subspecies != "Unidentified"]), 
                               collapse = ", "),
    Earliest = min(year, na.rm = TRUE),
    Latest = max(year, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(Records)) %>%
  slice_head(n = 20)

kable(location_stats, format = "html", 
      caption = "Table S2: Top 20 Locations with Complete Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, font_size = 12)
```


## 8.2 Session Information
```{r SessInf, include=TRUE}
cat("R version:", R.version$version.string, "\n")
cat("Platform:", R.version$platform, "\n")
cat("Date analyzed:", format(Sys.Date(), "%Y-%m-%d"), "\n\n")
cat("Key packages:\n")
session_info <- sessionInfo()
for (pkg in c("dplyr", "ggplot2", "sf", "leaflet", "plotly")) {
  if (pkg %in% names(session_info$otherPkgs)) {
    ver <- session_info$otherPkgs[[pkg]]$Version
    cat(paste0("  - ", pkg, ": ", ver, "\n"))
  }
}
```

